---
title: "DATA 606 Data Project Proposal"
author: "Adam Gersowitz"
output: pdf_document
---

### Data Preparation

```{r setup, echo=TRUE, results='hide', warning=FALSE, message=FALSE}
library(nasaweather)
library(sqldf)
nrow(atmos)
length(atmos)
head(atmos,10)

# I will create a dataframe of the information that could impact the temperature as well as the progression of time

temp <- atmos[c(1:6)]

# I will also create dataframes that store average information on temperature by date and location
temp<-sqldf("select a.*, lat||'-'||long ll from temp a")

meantemp_coord <-sqldf("select avg(temp) mean_temp,avg(surftemp) mean_surftemp, lat, long from temp group by lat, long order by lat,long")

meantemp_date  <-sqldf("select avg(temp) mean_temp,avg(surftemp) mean_surftemp, month, year from temp group by month, year order by month asc, year asc")

meantemp_year  <-sqldf("select avg(temp) mean_temp,avg(surftemp) mean_surftemp,  year from temp group by  year order by  year asc")

meantemp_coordyear  <-sqldf("select avg(temp) mean_temp,avg(surftemp) mean_surftemp, lat,long,year,ll from temp group by lat,long, year,ll order by year asc")

#Lastly I'll create a dataframe that captures the percent change from month to month

temp_percent_change  <-sqldf(" select tm.lat, tm.long, tm.month, tm.year, tm.temp last_month_temp, tm.surftemp, last_month_surftemp, (temp-last_month_temp)/last_month_temp*100 as chng_temp,
                                (surftemp-last_month_surftemp)/last_month_surftemp*100 as chng_surftemp
                                from
                              (select temp,surftemp, month, year, lat, long 
                              from temp) tm
                             join
                             (select temp as last_month_temp, surftemp as last_month_surftemp, lat, long, 
                            case when month = 12 then year-1 else year end year,
                            case when month = 12 then 1 else month+1 end month from temp) lm on
                             lm.lat = tm.lat and lm.long = tm.long and lm.year = tm.year and lm.month = tm.month
                             order by  tm.lat, tm.long
                             
                             ")

head(meantemp_coordyear,10)

```


### Research question 

Has the average monthly temperature changed at the same rate across different coordinates in Central America?



### Cases 

Each case represents an atmospheric weather reading of coordinates in Central America. There 41,472 cases in the data set.




### Data collection 

The data was collected by NASA and is data from the 2006 ASA data expo. 



### Type of study 

This is an observational study.



### Data Source 

The data can be found at http://stat-computing.org/dataexpo/2006/. For this project the data was pulled from the nasaweather package. (https://blog.rstudio.com/2014/07/23/new-data-packages/)




### Dependent Variable

The response variables are the temperature and surface temperatures which are numerical.



### Independent Variable

Teh explanatory variable is the cooridnates (laititude and longitude) which is categorical.



### Relevant summary statistics 

**Provide summary statistics for each the variables. Also include appropriate visualizations related to your research question (e.g. scatter plot, boxplots, etc). This step requires the use of R, hence a code chunk is provided below. Insert more code chunks as needed.**

```{r}

summary(temp_percent_change$chng_temp)

```
```{r}
#install.packages("Hmisc")

summary(temp_percent_change$chng_surftemp)
```

```{r}
library(ggplot2)
ggplot(data = meantemp_year , aes(x = year, y = mean_temp)) + 
     geom_line() +
     geom_point() +
     scale_x_discrete(breaks = meantemp_year$year, labels = meantemp_year$year)

```